A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-09-29 bioc-workflow.Rnw
move to Rnw, update license to CC-BY

\section{Introduction}\label{introduction}

@@

The main package to analyse protein localisation data is
\texttt{r Biocpkg("pRoloc")}, which offers a set of dedicated functions
for the analysis of such data.
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}} itself
relies on \texttt{r Biocpkg("MSnbase")} to manipulate and process
quantitative proteomics data. Many other packages are used by
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}} for
clustering, classification and visualisation.

In this workflow, we will describe how to prepare the spatial proteomics
data starting from a spreadsheet containing quantitative mass
spectrometry data. We will focus on a recent pluripotent mouse embryonic
stem cells experiment (A. Christoforou et al. 2016). Additional
annotated and pre-formatted datasets from various species are readily
available in the
\emph{\href{http://bioconductor.org/packages/pRolocdata}{pRolocdata}}
package.

Installation of Bioconductor package is documented in details on the
\href{http://bioconductor.org/install/\#install-bioconductor-packages}{Bioconductor
installation help page}. This procedure is also applicable to any
packages, from \href{https://cran.r-project.org/}{CRAN} as well as
GitHub. Once a package has been installed, it needs to be loaded for its
functionality to become available in the R session; this is done with
the \texttt{library} function e.g. to load the
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}} one would
type \texttt{library("pRoloc")} after installation.

If you have questions about this workflow in particular, or about other
Bioconductor packages in general, they are best asked on the
\href{https://support.bioconductor.org/}{Bioconductor support site}
following the
\href{http://www.bioconductor.org/help/support/posting-guide/}{posting
guidelines}. Questions can be tagged with specific package names or
keywords. For more general information about mass spectrometry and
proteomics, the readers are invited to read the
\emph{\href{http://bioconductor.org/packages/RforProteomics}{RforProteomics}}
package vignettes and associated papers (L. Gatto and Christoforou 2014;
Gatto et al. 2015).

\section{Reading and processing spatial proteomics
data}\label{reading-and-processing-spatial-proteomics-data}

\subsection{The use-case: predicting sub-cellular localisation in
pluripotent embryonic mouse stem
cells}\label{the-use-case-predicting-sub-cellular-localisation-in-pluripotent-embryonic-mouse-stem-cells}

As a use-case, we analyse a recent high-throughput spatial proteomics
dataset from pluripotent mouse embryonic stem cells (E14TG2a) (A.
Christoforou et al. 2016). The data was generated using hyperplexed
LOPIT (hyperLOPIT), a state-of-the-art method relying on improved
sub-cellular fractionation and more accurate quantitation, leading to
more reliable classification of protein localisation across the whole
sub-cellular space. The method uses an elaborate sub-cellular
fractionation scheme, enabled by the use of Tandem Mass Tag (TMT)
{[}Thompson:2003{]} 10-plex and application of the MS data acquisition
technique named synchronous precursor selection MS\textsuperscript{3}
(SPS-MS\textsuperscript{3}) (McAlister et al. 2014), for TMT
quantification with high accuracy and precision. Three biological
replicates were generated from the E14TG2a experiment, the first was to
target low density fractions and the second and third were to emphasis
separation of the denser organelles. The intersect of replicates 1 and 2
was treated as a 20-plex dataset for the analysis. As discussed in the
manuscript (A. Christoforou et al. 2016) as it has been shown that
combining replicates across from different gradients can increase
spatial resolution (M. W. B. Trotter et al. 2010). The combination of
replicates resulted in 5032 proteins common in both experiments.

These, as well as many other data are directly available as properly
structured and annotated computational object from the
\emph{\href{http://bioconductor.org/packages/pRolocdata}{pRolocdata}}
experiment package. In this workflow, we will start with a description
of how to generate these ad hoc objects starting from an arbitrary
spreadsheets, as produced by many popular third-party applications.

While we focus here on a LOPIT-type dataset, these analyses are relevant
for any quantitative spatial proteomics data, irrespective of the
fractionation or quantitation (i.e.~labelled or label-free) methods.

\subsection{The infrastructure:
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}} and
\emph{\href{http://bioconductor.org/packages/MSnbase}{MSnbase}} in
Bioconductor}\label{the-infrastructure-proloc-and-msnbase-in-bioconductor}

To make use of the full functionality of the
\emph{\href{http://bioconductor.org/packages/pRoloc}{pRoloc}} software
one needs to import their data into R and prepare them as an
\texttt{MSnSet}. The \texttt{MSnSet} is a dedicated data structure for
the efficient manipulation and processing of mass spectrometry and
proteomics data in R. Figure 1 illustrates a simplified view of the
\texttt{MSnSet} structure; there exists 3 key sub-parts (termed slots)
to such a data object: (1) the \texttt{exprs} slot for storing the
quantitation data, (2) the \texttt{fData} slot for storing the feature
meta-data, and finally (3) the \texttt{pData} slot for storing the
sample meta-data.

\begin{figure}[htbp]
\centering
\includegraphics{./Figures/msnset.png}
\caption{Simplified representation of the \texttt{MSnSet} data structure
(reproduced with permission from the
\emph{\href{http://bioconductor.org/packages/MSnbase}{MSnbase}}
vignette)}
\end{figure}

\subsection{Importing data}\label{importing-data}

There are a number of ways to import quantitation data and create an
\texttt{MSnSet} instance. All methods are described in the
\emph{\href{http://bioconductor.org/packages/MSnbase}{MSnbase}}
\href{http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-io.pdf}{input/output
capabilities vignette}. One suggested simple method is to use the
function \texttt{readMSnSet2} in
\emph{\href{http://bioconductor.org/packages/MSnbase}{MSnbase}}. The
function takes a single spreadsheet as input and extracts the columns
containing the quantitation data, as identified by the argument
\texttt{ecol}, to create the expression data, while the other columns in
the spreadsheet are appended to the feature meta-data slot. By example,
in the code chunk below we read in the \texttt{csv} spreadsheet
containing the quantitation data from the intersect of replicates 1 and
2 of the mouse map (A. Christoforou et al. 2016), using the
\texttt{readMSnSet2} function. The data is as available online with the
manuscript (see tab 2 of the \texttt{xlsx} supplementary data set 1 in
(A. Christoforou et al. 2016), which should be exported as a text-based
spreadsheet). It is also available as a \texttt{csv} in the Bioconductor
\texttt{r Biocexptpkg("pRolocdata")} data package.

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"MSnbase"}\NormalTok{)}
\NormalTok{extdatadir <-}\StringTok{ }\KeywordTok{system.file}\NormalTok{(}\StringTok{"extdata"}\NormalTok{, }\DataTypeTok{package =} \StringTok{"pRolocdata"}\NormalTok{)}
\NormalTok{csvfile <-}\StringTok{ }\KeywordTok{dir}\NormalTok{(extdatadir, }\DataTypeTok{full.names =} \OtherTok{TRUE}\NormalTok{,}
          \DataTypeTok{pattern =} \StringTok{"hyperLOPIT-SIData-ms3-rep12-intersect.csv"}\NormalTok{)}
\KeywordTok{basename}\NormalTok{(csvfile)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "hyperLOPIT-SIData-ms3-rep12-intersect.csv.gz"
\end{verbatim}

@@

\begin{figure}[htbp]
\centering
\includegraphics{./Figures/spreadsheet-screenshot.png}
\caption{A screenshot of the data in the spreadsheet.}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{getEcols}\NormalTok{(csvfile, }\DataTypeTok{split =} \StringTok{","}\NormalTok{, }\DataTypeTok{n =} \DecValTok{2}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  [1] ""                                 
##  [2] ""                                 
##  [3] ""                                 
##  [4] "Experiment 1"                     
##  [5] "Experiment 2"                     
##  [6] "Experiment 1"                     
##  [7] "Experiment 2"                     
##  [8] "126"                              
##  [9] "127N"                             
## [10] "127C"                             
## [11] "128N"                             
## [12] "128C"                             
## [13] "129N"                             
## [14] "129C"                             
## [15] "130N"                             
## [16] "130C"                             
## [17] "131"                              
## [18] "126"                              
## [19] "127N"                             
## [20] "127C"                             
## [21] "128N"                             
## [22] "128C"                             
## [23] "129N"                             
## [24] "129C"                             
## [25] "130N"                             
## [26] "130C"                             
## [27] "131"                              
## [28] "phenoDisco Input"                 
## [29] "phenoDisco Output"                
## [30] "Curated phenoDisco Output"        
## [31] "SVM marker set"                   
## [32] "SVM classification"               
## [33] "SVM score"                        
## [34] "SVM classification (top quartile)"
## [35] "Final Localization Assignment"    
## [36] "First localization evidence?"     
## [37] "Curated Organelles"               
## [38] "Cytoskeletal Components"          
## [39] "Trafficking Proteins"             
## [40] "Protein Complexes"                
## [41] "Signaling Cascades"               
## [42] "Oct4 Interactome"                 
## [43] "Nanog Interactome"                
## [44] "Sox2 Interactome"                 
## [45] "Cell Surface Proteins"
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hl <-}\StringTok{ }\KeywordTok{readMSnSet2}\NormalTok{(csvfile, }\DataTypeTok{ecol =} \DecValTok{8}\NormalTok{:}\DecValTok{27}\NormalTok{, }\DataTypeTok{fnames =} \DecValTok{1}\NormalTok{, }\DataTypeTok{skip =} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Below, we display a short summary of the data. The data contains 5032
proteins/features common across the 2 biological replicates for the
respective 2 x 10-plex reporter tags (20 columns/samples), along with
associated feature meta-data such as protein markers, protein
description, number of quantified peptides etc (see below).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hl}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData: none
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: X X.1 ... Cell.Surface.Proteins (25 total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
##  MSnbase version: 1.99.2
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{exprs}\NormalTok{(hl)[}\DecValTok{1}\NormalTok{:}\DecValTok{5}\NormalTok{, ]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##           X126 X127N X127C X128N X128C X129N X129C X130N X130C  X131
## Q9JHU4   0.028 0.034 0.024 0.014 0.026 0.045 0.107 0.341 0.059 0.321
## Q9QXS1-3 0.039 0.134 0.095 0.053 0.084 0.121 0.107 0.128 0.122 0.117
## Q9ERU9   0.021 0.013 0.014 0.009 0.024 0.054 0.116 0.257 0.209 0.284
## P26039   0.120 0.255 0.148 0.091 0.135 0.095 0.041 0.057 0.014 0.043
## Q8BTM8   0.055 0.139 0.078 0.050 0.077 0.098 0.093 0.171 0.079 0.160
##          X126.1 X127N.1 X127C.1 X128N.1 X128C.1 X129N.1 X129C.1 X130N.1
## Q9JHU4    0.037   0.064   0.058   0.059   0.067   0.078   0.140   0.208
## Q9QXS1-3  0.033   0.073   0.074   0.062   0.081   0.142   0.190   0.069
## Q9ERU9    0.026   0.017   0.023   0.029   0.039   0.071   0.105   0.171
## P26039    0.111   0.181   0.141   0.144   0.152   0.119   0.075   0.028
## Q8BTM8    0.062   0.108   0.091   0.086   0.099   0.111   0.117   0.095
##          X130C.1 X131.1
## Q9JHU4     0.141  0.147
## Q9QXS1-3   0.151  0.125
## Q9ERU9     0.304  0.215
## P26039     0.017  0.033
## Q8BTM8     0.144  0.087
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{exprs}\NormalTok{(hl)[}\KeywordTok{c}\NormalTok{(}\StringTok{"Q9ERU9"}\NormalTok{, }\StringTok{"Q9Z2R6"}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\StringTok{"X126"}\NormalTok{, }\StringTok{"X131.1"}\NormalTok{)]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##         X126 X131.1
## Q9ERU9 0.021  0.215
## Q9Z2R6 0.563  0.000
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{fvarLabels}\NormalTok{(hl)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  [1] "X"                                
##  [2] "X.1"                              
##  [3] "X.2"                              
##  [4] "Experiment.1"                     
##  [5] "Experiment.2"                     
##  [6] "Experiment.1.1"                   
##  [7] "Experiment.2.1"                   
##  [8] "phenoDisco.Input"                 
##  [9] "phenoDisco.Output"                
## [10] "Curated.phenoDisco.Output"        
## [11] "SVM.marker.set"                   
## [12] "SVM.classification"               
## [13] "SVM.score"                        
## [14] "SVM.classification..top.quartile."
## [15] "Final.Localization.Assignment"    
## [16] "First.localization.evidence."     
## [17] "Curated.Organelles"               
## [18] "Cytoskeletal.Components"          
## [19] "Trafficking.Proteins"             
## [20] "Protein.Complexes"                
## [21] "Signaling.Cascades"               
## [22] "Oct4.Interactome"                 
## [23] "Nanog.Interactome"                
## [24] "Sox2.Interactome"                 
## [25] "Cell.Surface.Proteins"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{fvarLabels}\NormalTok{(hl)[}\DecValTok{1}\NormalTok{:}\DecValTok{3}\NormalTok{] <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"uniprot.accession"}\NormalTok{, }\StringTok{"uniprot.id"}\NormalTok{, }\StringTok{"description"}\NormalTok{)}
\KeywordTok{fvarLabels}\NormalTok{(hl)[}\DecValTok{4}\NormalTok{:}\DecValTok{6}\NormalTok{] <-}\StringTok{ }\KeywordTok{paste0}\NormalTok{(}\StringTok{"peptides.expt"}\NormalTok{, }\DecValTok{1}\NormalTok{:}\DecValTok{3}\NormalTok{)}
\KeywordTok{fData}\NormalTok{(hl)[}\DecValTok{1}\NormalTok{:}\DecValTok{4}\NormalTok{, }\DecValTok{1}\NormalTok{:}\DecValTok{6}\NormalTok{]}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##          uniprot.accession  uniprot.id
## Q9JHU4              Q9JHU4 DYHC1_MOUSE
## Q9QXS1-3          Q9QXS1-3  PLEC_MOUSE
## Q9ERU9              Q9ERU9  RBP2_MOUSE
## P26039              P26039  TLN1_MOUSE
##                                                                                                                                 description
## Q9JHU4                                              Cytoplasmic dynein 1 heavy chain 1 OS=Mus musculus GN=Dync1h1 PE=1 SV=2 - [DYHC1_MOUSE]
## Q9QXS1-3 Isoform PLEC-1 of Plectin OS=Mus musculus GN=Plec - [PLEC_MOUSE]|Isoform PLEC-1A of Plectin OS=Mus musculus GN=Plec - [PLEC_MOUSE]
## Q9ERU9                                                     E3 SUMO-protein ligase RanBP2 OS=Mus musculus GN=Ranbp2 PE=1 SV=2 - [RBP2_MOUSE]
## P26039                                                                             Talin-1 OS=Mus musculus GN=Tln1 PE=1 SV=2 - [TLN1_MOUSE]
##          peptides.expt1 peptides.expt2 peptides.expt3
## Q9JHU4              175            166            322
## Q9QXS1-3            123            150            174
## Q9ERU9              101             90            181
## P26039              101             94            167
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{pData}\NormalTok{(hl)$Replicate <-}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\NormalTok{:}\DecValTok{2}\NormalTok{, }\DataTypeTok{each =} \DecValTok{10}\NormalTok{)}
\KeywordTok{pData}\NormalTok{(hl)$Tag <-}\StringTok{ }\KeywordTok{sub}\NormalTok{(}\StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{.1$"}\NormalTok{, }\StringTok{""}\NormalTok{, }\KeywordTok{sub}\NormalTok{(}\StringTok{"^X"}\NormalTok{, }\StringTok{""}\NormalTok{, }\KeywordTok{sampleNames}\NormalTok{(hl)))}
\KeywordTok{pData}\NormalTok{(hl)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##         Replicate  Tag
## X126            1  126
## X127N           1 127N
## X127C           1 127C
## X128N           1 128N
## X128C           1 128C
## X129N           1 129N
## X129C           1 129C
## X130N           1 130N
## X130C           1 130C
## X131            1  131
## X126.1          2  126
## X127N.1         2 127N
## X127C.1         2 127C
## X128N.1         2 128N
## X128C.1         2 128C
## X129N.1         2 129N
## X129C.1         2 129C
## X130N.1         2 130N
## X130C.1         2 130C
## X131.1          2  131
\end{verbatim}

Throughout this workflow we refer to the different columns that are
found in the \texttt{exprs} (expression data) slot as channels (short
for TMT channels). In the frame of LOPIT and hyperLOPIT these channels
consititue the relative abundance of each protein (along the rows) in
the channel of interest. Each TMT channel originates from fractions
collected from the density gradient, or a set of pooled fractions or may
be a sample originating from an alternative preparation e.g.~such as
from the chromatin enrichment performed in Christoforou et al (A.
Christoforou et al. 2016). Information about which gradient fractions
were used for which tag should also be stored in the sample meta-data
\texttt{pData} slot.

The sample meta-data that is distributed with the
\emph{\href{http://bioconductor.org/packages/pRolocdata}{pRolocdata}}
package for Christoforou's hyperLOPIT experiment and (as above) the
quantitation data file, are located in the \texttt{extdata} in the
\emph{\href{http://bioconductor.org/packages/pRolocdata}{pRolocdata}}
package on the hard drive.

In the code chunk below we again use the \texttt{dir} function to locate
the filepath to the meta-data \texttt{csv} file and then read it into R
using \texttt{read.csv}. We then append the meta-data to the
\texttt{pData} slot. Information about the gradient fractions used and
the associated subcellular fraction densities in \% w/v Iodixanol for
each replicate are stored here.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{expinfo <-}\StringTok{ }\KeywordTok{dir}\NormalTok{(extdatadir, }\DataTypeTok{full.names =} \OtherTok{TRUE}\NormalTok{,}
               \DataTypeTok{pattern =} \StringTok{"hyperLOPIT-SIData-fraction-info.csv"}\NormalTok{)}

\NormalTok{fracinfo <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(expinfo, }\DataTypeTok{row.names=}\DecValTok{1}\NormalTok{, }\DataTypeTok{skip =} \DecValTok{2}\NormalTok{, }
                     \DataTypeTok{header =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{stringsAsFactors =} \OtherTok{FALSE}\NormalTok{)}

\KeywordTok{pData}\NormalTok{(hl)$Gradient.Fraction <-}\StringTok{ }\KeywordTok{c}\NormalTok{(fracinfo[, }\DecValTok{1}\NormalTok{], fracinfo[, }\DecValTok{2}\NormalTok{])}
\KeywordTok{pData}\NormalTok{(hl)$Iodixonal.Density <-}\StringTok{ }\KeywordTok{c}\NormalTok{(fracinfo[, }\DecValTok{4}\NormalTok{], fracinfo[, }\DecValTok{5}\NormalTok{])}
\KeywordTok{pData}\NormalTok{(hl)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##         Replicate  Tag Gradient.Fraction Iodixonal.Density
## X126            1  126           Cytosol               0.0
## X127N           1 127N   1 to 6 (pooled)               6.0
## X127C           1 127C   8 to 9 (pooled)              11.0
## X128N           1 128N 10 to 11 (pooled)              13.3
## X128C           1 128C                12              14.6
## X129N           1 129N                14              17.4
## X129C           1 129C                16              20.1
## X130N           1 130N                18              26.8
## X130C           1 130C         Chromatin                NA
## X131            1  131                19              34.5
## X126.1          2  126           Cytosol               0.0
## X127N.1         2 127N   1 to 6 (pooled)               5.2
## X127C.1         2 127C   7 to 9 (pooled)              10.0
## X128N.1         2 128N 10 to 11 (pooled)              12.5
## X128C.1         2 128C                12              14.0
## X129N.1         2 129N 14 to 15 (pooled)              17.3
## X129C.1         2 129C                17              20.9
## X130N.1         2 130N 18 to 19 (pooled)              24.7
## X130C.1         2 130C         Chromatin                NA
## X131.1          2  131                20              31.9
\end{verbatim}

\subsection{Data processing}\label{data-processing}

\subsubsection{Normalisation}\label{normalisation}

There are two aspects related to data normalisation that are relevant to
spatial proteomics data processing. The first one focuses on reducing
purely technical variation between channels without affecting biological
variability (i.e.~the shape of the quantitative profiles). This
normalisation will depend on the underlying quantitative technology and
the experimental design, and will not be addressed in this workflow. The
second aspect, and more specific to spatial proteomics data, is scaling
all the organelle-specific profiles into a same intensity interval
(typically 0 and 1) by, for example, diving each intensity by the sum of
the intensities for that quantitative feature. This is not necessary in
this example as the intensities for each replicate have already been
re-scaled to 1 in Proteome Discoverer. However, if one wanted to do this
they would execute the \texttt{normalise} function as demonstrated in
the below code chunk.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hl <-}\StringTok{ }\KeywordTok{normalise}\NormalTok{(hl, }\DataTypeTok{method =} \StringTok{"sum"}\NormalTok{) }
\end{Highlighting}
\end{Shaded}

@@

As previously mentioned, before combination, the two replicates in the
\texttt{hl} data that we read into R were separately normalised by sum
(i.e.~to 1) across the 10 channels for each replicate respectively. We
can verify this by summing each rows for each replicate:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(}\KeywordTok{rowSums}\NormalTok{(}\KeywordTok{exprs}\NormalTok{(hl[, hl$replicate ==}\StringTok{ }\DecValTok{1}\NormalTok{])))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
##       0       0       0       0       0       0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{summary}\NormalTok{(}\KeywordTok{rowSums}\NormalTok{(}\KeywordTok{exprs}\NormalTok{(hl[, hl$replicate ==}\StringTok{ }\DecValTok{2}\NormalTok{])))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
##       0       0       0       0       0       0
\end{verbatim}

@@

\subsubsection{Combining acquisitions}\label{combining-acquisitions}

>>>>

<<env, echo=FALSE, warning=FALSE>>=
library("BiocStyle")
suppressPackageStartupMessages(library("knitr"))
suppressPackageStartupMessages(library("gridExtra"))
suppressPackageStartupMessages(library("MSnbase"))
suppressPackageStartupMessages(library("pRoloc"))
suppressPackageStartupMessages(library("pRolocdata"))
cache <- TRUE
@

\section*{Introduction}

@@

The main package to analyse protein localisation data is
\Biocpkg{pRoloc}, which offers a set of dedicated functions for the
analysis of such data. \Biocpkg{pRoloc} itself relies on
\Biocpkg{MSnbase} to manipulate and process quantitative
proteomics data. Many other packages are used by \Biocpkg{pRoloc}
for clustering, classification and visualisation.

In this workflow, we will describe how to prepare the spatial
proteomics data starting from a spreadsheet containing quantitative
mass spectrometry data. We will focus on a recent pluripotent mouse
embryonic stem cells experiment \cite{hyper}. Additional annotated and
pre-formatted datasets from various species are readily available in
the \Biocexptpkg{pRolocdata} package.

Installation of Bioconductor package is documented in details on the
\href{http://bioconductor.org/install/#install-bioconductor-packages}{Bioconductor
  installation help page}. This procedure is also applicable to any
packages, from \href{https://cran.r-project.org/}{CRAN} as well as
GitHub. Once a package has been installed, it needs to be loaded for
its functionality to become available in the R session; this is done
with the \texttt{library} function e.g.  to load the
\Biocpkg{pRoloc} one would type \texttt{library("pRoloc")}
after installation.

If you have questions about this workflow in particular, or about
other Bioconductor packages in general, they are best asked on the
\href{https://support.bioconductor.org/}{Bioconductor support site}
following the
\href{http://www.bioconductor.org/help/support/posting-guide/}{posting
  guidelines}. Questions can be tagged with specific package names or
keywords. For more general information about mass spectrometry and
proteomics, the readers are invited to read the
\Biocexptpkg{"RforProteomics"} package vignettes and associated
papers \cite{Gatto:2014,Gatto:2015}.

\section*{Reading and processing spatial proteomics data}

\subsection*{The use-case: predicting sub-cellular localisation in pluripotent embryonic mouse stem cells}

As a use-case, we analyse a recent high-throughput spatial proteomics
dataset from pluripotent mouse embryonic stem cells (E14TG2a)
\cite{hyper}. The data was generated using hyperplexed LOPIT
(hyperLOPIT), a state-of-the-art method relying on improved
sub-cellular fractionation and more accurate quantitation, leading to
more reliable classification of protein localisation across the whole
sub-cellular space. The method uses an elaborate sub-cellular
fractionation scheme, enabled by the use of Tandem Mass Tag (TMT)
\cite{Thompson:2003} 10-plex and application of the MS data
acquisition technique named synchronous precursor selection MS$^3$
(SPS-MS$^3$) \cite{McAlister:2014}, for TMT quantification with high
accuracy and precision. Three biological replicates were generated
from the E14TG2a experiment, the first was to target low density
fractions and the second and third were to emphasis separation of the
denser organelles.  The intersect of replicates 1 and 2 was treated as
a 20-plex dataset for the analysis.  As discussed in the manuscript
\cite{hyper} as it has been shown that combining replicates across
from different gradients can increase spatial resolution
\cite{Trotter:2010}. The combination of replicates resulted in 5032
proteins common in both experiments.

These, as well as many other data are directly available as properly
structured and annotated computational object from the 
\Biocexptpkg{pRolocdata} experiment package. In this workflow, we
will start with a description of how to generate these ad hoc objects
starting from an arbitrary spreadsheets, as produced by many popular
third-party applications.

While we focus here on a LOPIT-type dataset, these analyses are
relevant for any quantitative spatial proteomics data, irrespective of
the fractionation or quantitation (i.e. labelled or label-free)
methods.

\subsection*{The infrastructure: \texttt{pRoloc} and \texttt{MSnbase} packages}

To make use of the full functionality of the \Biocpkg{pRoloc}
software one needs to import their data into R and prepare them as an
\texttt{MSnSet}. The \texttt{MSnSet} is a dedicated data structure for the efficient
manipulation and processing of mass spectrometry and proteomics data
in R. Figure 1 illustrates a simplified view of the \texttt{MSnSet}
structure; there exists 3 key sub-parts (termed slots) to such a data
object: (1) the \texttt{exprs} slot for storing the quantitation data, (2)
the \texttt{fData} slot for storing the feature meta-data, and finally (3)
the \texttt{pData} slot for storing the sample meta-data.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.45\textwidth]{./Figures/msnset.png}
  \caption{Simplified representation of the \texttt{MSnSet} data
    structure (reproduced with permission from the \Biocpkg{MSnbase}
    vignette)}
  \label{fig:msnset}
\end{figure}

\subsection*{Importing data}

There are a number of ways to import quantitation data and create an
\texttt{MSnSet} instance. All methods are described in the
\Biocpkg{MSnbase}
\href{http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-io.pdf}{input/output
  capabilities vignette}. One suggested simple method is to use the
function \texttt{readMSnSet2} in \Biocpkg{MSnbase}. The function
takes a single spreadsheet as input and extracts the columns
containing the quantitation data, as identified by the argument
\texttt{ecol}, to create the expression data, while the other columns
in the spreadsheet are appended to the feature meta-data slot.  By
example, in the code chunk below we read in the \texttt{csv}
spreadsheet containing the quantitation data from the intersect of
replicates 1 and 2 of the mouse map \cite{hyper}, using the
\texttt{readMSnSet2} function. The data is as available online with
the manuscript (see tab 2 of the \texttt{xlsx} supplementary data set
1 in \cite{hyper}, which should be exported as a text-based
spreadsheet). It is also available as a \texttt{csv} in the
Bioconductor \Biocexptpkg{pRolocdata} data package.

@@

<<getFilename, cache = cache>>=
library("MSnbase")
extdatadir <- system.file("extdata", package = "pRolocdata")
csvfile <- dir(extdatadir, full.names = TRUE,
          pattern = "hyperLOPIT-SIData-ms3-rep12-intersect.csv")
basename(csvfile)
@

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.65\textwidth]{./Figures/spreadsheet-screenshot.png}
  \caption{A screenshot of the data in the spreadsheet.}
  \label{fig:spreadsheet}
\end{figure}

@@

<<getEcols>>=
getEcols(csvfile, split = ",", n = 2)
@

@@

<<readMSnSet2>>=
hl <- readMSnSet2(csvfile, ecol = 8:27, fnames = 1, skip = 1)
@

Below, we display a short summary of the data. The data contains 
\Sexpr{nrow(hl)} proteins/features common across the 2 biological replicates
for the respective 2 x 10-plex reporter tags (\Sexpr{ncol(hl)}
columns/samples), along with associated feature meta-data such as
protein markers, protein description, number of quantified peptides
etc (see below).

<<showhl>>=
hl
@

@@

<<assesshl>>=
exprs(hl)[1:5, ]
exprs(hl)[c("Q9ERU9", "Q9Z2R6"), c("X126", "X131.1")]
@

@@

<<removefData>>=
fvarLabels(hl)
fvarLabels(hl)[1:3] <- c("uniprot.accession", "uniprot.id", "description")
fvarLabels(hl)[4:6] <- paste0("peptides.expt", 1:3)
fData(hl)[1:4, 1:6]
@

@@

<<pdata>>=
pData(hl)$Replicate <- rep(1:2, each = 10)
pData(hl)$Tag <- sub("\\.1$", "", sub("^X", "", sampleNames(hl)))
pData(hl)
@

Throughout this workflow we refer to the different columns that
are found in the \texttt{exprs} (expression data) slot as channels (short for TMT channels).
In the frame of LOPIT and hyperLOPIT these channels consititue the
relative abundance of each protein (along the rows) in the channel of
interest. Each TMT channel originates from fractions collected from
the density gradient, or a set of pooled fractions or may be a sample
originating from an alternative preparation e.g. such as from the
chromatin enrichment performed in Christoforou et al \cite{hyper}.
Information about which gradient fractions were used for which tag
should also be stored in the sample meta-data \texttt{pData} slot.

The sample meta-data that is distributed with the
\Biocexptpkg{pRolocdata} package for Christoforou's
hyperLOPIT experiment and (as above) the quantitation data file, are
located in the \texttt{extdata} in the \Biocexptpkg{pRolocdata}
package on the hard drive.

In the code chunk below we again use the
\texttt{dir} function to locate the filepath to the meta-data \texttt{csv} file and
then read it into R using \texttt{read.csv}. We then append the meta-data to
the \texttt{pData} slot.  Information about the gradient fractions used and
the associated subcellular fraction densities in % w/v Iodixanol for
each replicate are stored here.

<<morepdata>>=
expinfo <- dir(extdatadir, full.names = TRUE,
               pattern = "hyperLOPIT-SIData-fraction-info.csv")

fracinfo <- read.csv(expinfo, row.names=1, skip = 2, 
                     header = FALSE, stringsAsFactors = FALSE)

pData(hl)$Gradient.Fraction <- c(fracinfo[, 1], fracinfo[, 2])
pData(hl)$Iodixonal.Density <- c(fracinfo[, 4], fracinfo[, 5])
pData(hl)
@

\subsection*{Data processing}

\subsubsection*{Normalisation}

There are two aspects related to data normalisation that are relevant
to spatial proteomics data processing. The first one focuses on
reducing purely technical variation between channels without
affecting biological variability (i.e. the shape of the quantitative
profiles). This normalisation will depend on the underlying
quantitative technology and the experimental design, and will not be
addressed in this workflow. The second aspect, and more specific to
spatial proteomics data, is scaling all the organelle-specific
profiles into a same intensity interval (typically 0 and 1) by, for
example, diving each intensity by the sum of the intensities for that
quantitative feature. This is not necessary in this example as the
intensities for each replicate have already been re-scaled to 1 in
Proteome Discoverer. However, if one wanted to do this they would
execute the \texttt{normalise} function as demonstrated in the below code
chunk.

<<normsum, eval = FALSE>>= 
hl <- normalise(hl, method = "sum") 
@

@@

As previously mentioned, before combination, the two replicates in the
\texttt{hl} data that we read into R were separately normalised by sum (i.e.
to 1) across the 10 channels for each replicate respectively. We can
verify this by summing each rows for each replicate:

<<normcheck>>=
summary(rowSums(exprs(hl[, hl$replicate == 1])))
summary(rowSums(exprs(hl[, hl$replicate == 2])))
@

@@

\subsubsection*{Combining acquisitions}

The spreadsheet that was used to create the \texttt{hl} MSnSet
included the two replicates within one .csv file.  We also provide
individual replicates in the \Biocexptpkg{pRolocdata}
package. Below, we show how to combine \texttt{MSnSet} objects and,
subsequently, how to filter and handle missing values. We start by
loading the \Biocexptpkg{pRolocdata} package and the
equivalent replicates using the \texttt{data} function.

<<reps4combine>>=
library("pRolocdata")
data(hyperLOPIT2015ms3r1)
data(hyperLOPIT2015ms3r2)
@
